---
title: "Final_project"
output: pdf_document
date: "2024-11-17"
---

## Exploratory Data Analysis  

```{r, warning=FALSE}
library(tidyverse)
library(glmnet) 
library(rstan) 
library(bayesplot)
library(gridExtra)
library(caret)


data <- read.csv("data/h1n1_vaccine_prediction.csv")
data[data == ""] <- NA
categorical_vars <- c("age_bracket", "qualification", "race", "sex", 
                      "income_level", "marital_status", "housing_status", 
                      "employment", "census_msa")
summary(data)
mode_func <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}
data_imputed <- data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(all_of(categorical_vars), ~ ifelse(is.na(.), mode_func(.), .)))
data_encoded <- data_imputed %>%
  mutate(
    income_level = as.numeric(factor(income_level, 
                                     levels = c("Below Poverty", "<= $75,000, Above Poverty", "> $75,000"), 
                                     ordered = TRUE)),
    age_bracket = as.numeric(factor(age_bracket, 
                                    levels = c("18 - 34 Years", "35 - 44 Years", "45 - 54 Years", "55 - 64 Years", "65+ Years"), 
                                    ordered = TRUE)),
    qualification = as.numeric(factor(qualification, 
                                      levels = c("< 12 Years", "12 Years", "Some College", "College Graduate"), 
                                      ordered = TRUE)),
    census_msa = as.numeric(factor(census_msa, 
                                   levels = c("Non-MSA", "MSA, Not Principle  City", "MSA, Principle City"), 
                                   ordered = TRUE)),
    across(setdiff(categorical_vars, c("income_level", "age_bracket", "qualification", "census_msa")), 
           ~ as.numeric(as.factor(.)))
  )

set.seed(123)
train_index <- createDataPartition(data_encoded$h1n1_vaccine, p = 0.7, list = FALSE)
train_data <- data_encoded[train_index, ]
test_data <- data_encoded[-train_index, ]

```


```{r}
important_vars <- c("h1n1_vaccine", "census_msa", "income_level", "race", "qualification")
data_imputed$income_level <- factor(data_imputed$income_level,levels = unique(data_imputed$income_level), labels = c("Poverty", "Mid Income", "High Income"))
data_imputed$census_msa <- factor(data_imputed$census_msa,levels = unique(data_imputed$census_msa), labels = c("MSA, non-p city", "MSA, p city", "Non-MSA"))
plots <- lapply(important_vars, function(var) {
  ggplot(data_imputed, aes_string(x = var)) +
    geom_bar(fill = "skyblue", color = "black") +
    labs(title = paste(var), x = var, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.title = element_text(size = 14, face = "bold"))
})
grid.arrange(grobs = plots, ncol = 3, nrow = 2)

```

```{r}
exploratory_vars <- c("census_msa", "income_level", "race", "qualification")
plots <- lapply(exploratory_vars, function(var) {
  data_encoded %>%
    group_by(.data[[var]]) %>%
    summarize(mean_vaccine = mean(h1n1_vaccine, na.rm = TRUE)) %>%
    ggplot(aes(x = .data[[var]], y = mean_vaccine)) +
    geom_point(size = 3) +
    geom_line(group = 1) +
    labs(
      x = var,
      y = "Proportion Vaccinated"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold")
    )
})

grid.arrange(grobs = plots, ncol = 2, nrow = 2)
```

The plots show how census_msa, income_level, race, and qualification influence H1N1 vaccination rates. Qualification displays the most linear and significant trend, with vaccination rates steadily increasing across education levels, suggesting a strong relationship between education and vaccine uptake. In contrast, income_level also shows an upward trend, while race and census_msa exhibit more moderate or non-linear patterns. The clear differences between education levels make qualification an ideal candidate for assigning distinct alphas in a Bayesian hierarchical structure.

```{r}
x_train <- as.matrix(train_data %>% select(-h1n1_vaccine, -unique_id, -qualification))
x_test <- as.matrix(test_data %>% select(-h1n1_vaccine, -unique_id, -qualification))
y_train <- train_data$h1n1_vaccine

feature_scores <- apply(x_train, 2, function(col) {
  test <- glm(y_train ~ col, family = "binomial") 
  summary(test)$coefficients[2, 4]         
})

selected_features <- which(feature_scores < 0.05)
x_selected <- x_train[, selected_features]

x_test_selected <- x_test[, selected_features]
selected_variable_names <- colnames(x_train)[selected_features]
print(selected_variable_names)

original_vars <- colnames(x_train)
removed_features <- setdiff(original_vars, selected_variable_names)
print(removed_features)

```

The lasso regression uses L1 penalty that forces some coefficients to become exactly zero, so effectively excluding those variables from the model. "wash_hands_frequently" is eliminated variables because they likely showed weak or no significant correlation with the likelihood of receiving the H1N1 vaccine compared to other variables.

```{r}
stan_data <- list(
  N = nrow(train_data),
  J = length(unique(train_data$qualification)),
  K = ncol(x_selected),
  y = y_train,
  x = x_selected,
  qualification = as.numeric(as.factor(train_data$qualification))
)

stan_model_code_vec <- "
data {
  int<lower=0> N;
  int<lower=0> J;
  int<lower=0> K;
  int<lower=0,upper=1> y[N];
  matrix[N, K] x;
  int<lower=1,upper=J> qualification[N];
}
parameters {
  vector[K] beta;
  vector[J] alpha;
  real mu_alpha;
  real<lower=0> sigma_alpha;
}
model {
  // Priors
  alpha ~ normal(mu_alpha, sigma_alpha);
  beta ~ normal(0, 3);
  mu_alpha ~ normal(0, 3);
  sigma_alpha ~ cauchy(0, 2.5);
  
  // Vectorized likelihood
  y ~ bernoulli_logit(alpha[qualification] + x * beta);
}
generated quantities {
  vector[N] theta;
  vector[N] log_lik;
  for (n in 1:N) {
    theta[n] = inv_logit(alpha[qualification[n]] + x[n] * beta);
    log_lik[n] = bernoulli_lpmf(y[n] | theta[n]); 
  }
}
"

stan_model_vec <- stan_model(model_code = stan_model_code_vec)
fit_final <- sampling(stan_model_vec, data = stan_data, iter = 2000, chains = 4, seed = 123, cores = 4)
```

```{r}
print(fit_final, pars = c("beta", "alpha", "mu_alpha", "sigma_alpha"))
```

```{r}
x_train_2 <- as.matrix(train_data %>% select(-h1n1_vaccine, -unique_id, -census_msa))
x_test_2 <- as.matrix(test_data %>% select(-h1n1_vaccine, -unique_id, -census_msa))
feature_scores <- apply(x_train_2, 2, function(col) {
  test <- glm(y_train ~ col, family = "binomial") 
  summary(test)$coefficients[2, 4]         
})

selected_features_2 <- which(feature_scores < 0.05)
x_selected_2 <- x_train_2[, selected_features_2]
x_test_selected_2 <-x_test_2[, selected_features_2]
selected_variable_names_2 <- colnames(x_train_2)[selected_features_2]
print(selected_variable_names_2)

original_vars_2 <- colnames(x_train_2)
removed_features_2 <- setdiff(original_vars_2, selected_variable_names_2)
print(removed_features_2)
```

```{r}
stan_data_2 <- list(
  N = nrow(train_data),
  J = length(unique(train_data$census_msa)),
  K = ncol(x_selected_2),
  y = y_train,
  x = x_selected_2,
  census = as.numeric(as.factor(train_data$census_msa))
)

stan_model_code_vec_2 <- "
data {
  int<lower=0> N;
  int<lower=0> J;
  int<lower=0> K;
  int<lower=0,upper=1> y[N];
  matrix[N, K] x;
  int<lower=1,upper=J> census[N];
}
parameters {
  vector[K] beta;
  vector[J] alpha;
  real mu_alpha;
  real<lower=0> sigma_alpha;
}
model {
  // Priors
  alpha ~ normal(mu_alpha, sigma_alpha);
  beta ~ normal(0, 3);
  mu_alpha ~ normal(0, 3);
  sigma_alpha ~ cauchy(0, 1.5);
  
  // Vectorized likelihood
  y ~ bernoulli_logit(alpha[census] + x * beta);
}
generated quantities {
  vector[N] theta;
  vector[N] log_lik;
  for (n in 1:N) {
    theta[n] = inv_logit(alpha[census[n]] + x[n] * beta);
    log_lik[n] = bernoulli_lpmf(y[n] | theta[n]); 
  }
}
"

stan_model_vec_2 <- stan_model(model_code = stan_model_code_vec_2)
fit_final_2 <- sampling(stan_model_vec_2, data = stan_data_2, iter = 3000, chains = 4, seed = 123, cores = 4)
```

```{r}
library("loo")
log_lik_qualification <- extract_log_lik(fit_final, parameter_name = "log_lik")
log_lik_census <- extract_log_lik(fit_final_2, parameter_name = "log_lik")
loo_qualification <- loo(log_lik_qualification)
loo_census <- loo(log_lik_census)
comparison <- loo_compare(loo_qualification, loo_census)
print(comparison)
```

```{r}
# Prepare Stan-compatible test data
posterior_alpha <- extract(fit_final_2, pars = "alpha")$alpha  # Group-level intercepts
posterior_beta <- extract(fit_final_2, pars = "beta")$beta
stan_data_test <- list(
  N = nrow(test_data),  # Number of test observations
  J = length(unique(train_data$census)),  # Number of census groups (from training data)
  K = ncol(x_test_selected_2),  # Number of predictors
  x = x_test_selected_2,  # Test feature matrix
  census = as.numeric(as.factor(test_data$census_msa))  # Census group IDs
)

# Compute linear predictors for all test observations
linear_predictors <- sapply(1:nrow(x_test_selected_2), function(n) {
  posterior_alpha[, stan_data_test$census[n]] + 
    as.numeric(matrix(x_test_selected_2[n, ], nrow = 1) %*% t(posterior_beta))
})

predicted_probabilities <- colMeans(1 / (1 + exp(-linear_predictors)))
test_data$predicted_probabilities <- predicted_probabilities


```

```{r}
library(pROC)

# Compute AUC
auc <- roc(test_data$h1n1_vaccine, test_data$predicted_probabilities)$auc
cat("AUC:", auc, "\n")

```




```{r}
theta_samples <- extract(fit_final)$theta  # Extract samples for theta

colnames(theta_samples) <- paste0("theta[", 1:ncol(theta_samples), "]")

theta_summary <- apply(theta_samples, 2, function(x) {
  c(mean = mean(x), 
    median = median(x), 
    lower_95 = quantile(x, 0.025), 
    upper_95 = quantile(x, 0.975))
})

theta_summary_df <- as.data.frame(t(theta_summary))

# Visualize first 5 theta parameters with mcmc_areas
mcmc_areas(as.matrix(theta_samples[, 1:5]),
           pars = colnames(theta_samples)[1:5],
           prob = 0.95)

```
