---
title: "Final_project_code"
output: pdf_document
date: "2024-12-10"
---

# Data Cleaning and Exploratory Data Analysis  

```{r, warning=FALSE}
library(tidyverse)
library(glmnet) 
library(rstan) 
library(bayesplot)
library(gridExtra)
library(caret)


data <- read.csv("data/h1n1_vaccine_prediction.csv")
data[data == ""] <- NA
categorical_vars <- c("age_bracket", "qualification", "race", "sex", "income_level", "marital_status", "housing_status", "employment", "census_msa")
summary(data)

mode_func <- function(x) {
  uniq_x <- unique(x)
  uniq_x[which.max(tabulate(match(x, uniq_x)))]
}

data_imputed <- data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  mutate(across(all_of(categorical_vars), ~ ifelse(is.na(.), mode_func(.), .)))


data_encoded <- data_imputed %>%
  mutate(
    income_level = as.numeric(factor(income_level, 
                                     levels = c("Below Poverty", "<= $75,000, Above Poverty", "> $75,000"), 
                                     ordered = TRUE)),
    age_bracket = as.numeric(factor(age_bracket, 
                                    levels = c("18 - 34 Years", "35 - 44 Years", "45 - 54 Years", "55 - 64 Years", "65+ Years"), 
                                    ordered = TRUE)),
    qualification = as.numeric(factor(qualification, 
                                      levels = c("< 12 Years", "12 Years", "Some College", "College Graduate"), 
                                      ordered = TRUE)),
    census_msa = as.numeric(factor(census_msa, 
                                   levels = c("Non-MSA", "MSA, Not Principle  City", "MSA, Principle City"), 
                                   ordered = TRUE)),
    across(setdiff(categorical_vars, c("income_level", "age_bracket", "qualification", "census_msa")), 
           ~ as.numeric(as.factor(.)))
  )

set.seed(123)
train_index <- createDataPartition(data_encoded$h1n1_vaccine, p = 0.7, list = FALSE)
train_data <- data_encoded[train_index, ]
test_data <- data_encoded[-train_index, ]
```


```{r}
important_vars <- c("h1n1_vaccine", "census_msa", "income_level", "race", "qualification")
data_imputed$income_level <- factor(data_imputed$income_level,levels = unique(data_imputed$income_level), labels = c("Poverty", "Mid Income", "High Income"))
data_imputed$census_msa <- factor(data_imputed$census_msa,levels = unique(data_imputed$census_msa), labels = c("MSA, non-p city", "MSA, p city", "Non-MSA"))
data_imputed$qualification <- factor(data_imputed$qualification,levels = unique(data_imputed$qualification), labels = c("< 12 Years", "12 Years", "Some College", "College Graduate"))
plots <- lapply(important_vars, function(var) {
  ggplot(data_imputed, aes_string(x = var)) +
    geom_bar(fill = "skyblue", color = "black") +
    labs(title = paste(var), x = var, y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), 
          plot.title = element_text(size = 14, face = "bold"))
})
grid.arrange(grobs = plots, ncol = 3, nrow = 2)
```

```{r}
exploratory_vars <- c("census_msa", "income_level", "race", "qualification")
plots <- lapply(exploratory_vars, function(var) {
  data_encoded %>%
    group_by(.data[[var]]) %>%
    summarize(mean_vaccine = mean(h1n1_vaccine, na.rm = TRUE)) %>%
    ggplot(aes(x = .data[[var]], y = mean_vaccine)) +
    geom_point(size = 3) +
    geom_line(group = 1) +
    labs(x = var,y = "Proportion Vaccinated") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      plot.title = element_text(size = 14, face = "bold")
    )
})

grid.arrange(grobs = plots, ncol = 2, nrow = 2)
```

The plots show how census_msa, income_level, race(unranked), and qualification influence H1N1 vaccination rates. Qualification displays the most linear and significant trend, with vaccination rates increasing across education levels. Income_level also shows an upward trend, while census_msa exhibit more moderate or non-linear patterns. The clear differences between education levels make qualification an ideal candidate for assigning distinct alphas in a Bayesian hierarchical structure.

# Feature selection

```{r}
library(glmnet)

x_train <- as.matrix(train_data %>% select(-h1n1_vaccine, -unique_id, -qualification))
x_test <- as.matrix(test_data %>% select(-h1n1_vaccine, -unique_id, -qualification))
y_train <- train_data$h1n1_vaccine

lasso_model <- cv.glmnet(
  x_train, y_train, 
  alpha = 1,                    
  family = "binomial",         
  type.measure = "auc",         
  nfolds = 5
)

selected_coefficients <- as.matrix(coef(lasso_model, s = "lambda.min"))
selected_features <- rownames(selected_coefficients)[selected_coefficients[, 1] != 0][-1] 
print("Selected Features:")
print(selected_features)

x_selected <- x_train[, selected_features]
x_test_selected <- x_test[, selected_features]

selected_variable_names <- selected_features
original_vars <- colnames(x_train)
removed_features <- setdiff(original_vars, selected_variable_names)
print("Removed Features:")
print(removed_features)
```
# RStan Model (Qualification)

```{r}
stan_data <- list(
  N = nrow(train_data),
  J = length(unique(train_data$qualification)),
  K = ncol(x_selected),
  y = y_train,
  x = x_selected,
  qualification = as.numeric(as.factor(train_data$qualification))
)

stan_model_qualification <- "
data {
  int<lower=0> N;   // Total number of rows in the dataset
  int<lower=0> J;   // Number of unique levels in the qualification variable
  int<lower=0> K;   // Number of selected predictors
  int<lower=0,upper=1> y[N];   // Target variable (whether vaccinated or not)
  matrix[N, K] x;   // Selected predictors for each individual
  int<lower=1,upper=J> qualification[N];   // Encoded education levels for each individual
}

parameters {
  vector[K] beta;   // Coefficients for predictors
  vector[J] alpha;   // Separate intercept for each education level
  real mu_alpha;    // Hyperparameter for the intercepts: Mean of the group-level intercepts
  real<lower=0> sigma_alpha;   // Also Hyperparameter for the intercepts: Standard deviation of group-level intercepts
}

model {
  alpha ~ normal(mu_alpha, sigma_alpha);   // Group-level Intercepts that follow a normal distribution
  beta ~ normal(0, 1);   // Informative normal prior for predictors
  mu_alpha ~ normal(0, 1);   // Prior for mean intercepts
  sigma_alpha ~ cauchy(0, 1.5);   // Prior for standard deviation of intercepts
  y ~ bernoulli_logit(alpha[qualification] + x * beta);   // Binary outcome y with the logistic regression
}

generated quantities {
  vector[N] theta;   // Predicted probabilities
  vector[N] log_lik;   // Log-likelihood for each observation
  for (n in 1:N) {
    theta[n] = inv_logit(alpha[qualification[n]] + x[n] * beta);
    log_lik[n] = bernoulli_lpmf(y[n] | theta[n]); 
  }
}
"

stan_model_qualification <- stan_model(model_code = stan_model_qualification)
fit_final <- sampling(stan_model_qualification, data = stan_data, iter = 2000, chains = 4, seed = 123, cores = 4)
```

```{r}
print(fit_final, pars = c("beta", "alpha", "mu_alpha", "sigma_alpha"))
```

# Feature selection & stan model for census_msa

```{r}
x_train_2 <- as.matrix(train_data %>% select(-h1n1_vaccine, -unique_id, -census_msa))
x_test_2 <- as.matrix(test_data %>% select(-h1n1_vaccine, -unique_id, -census_msa))

lasso_model_2 <- cv.glmnet(
  x_train_2, y_train, 
  alpha = 1,                   
  family = "binomial",          
  type.measure = "auc",         
  nfolds = 5                  
)

selected_coefficients_2 <- as.matrix(coef(lasso_model_2, s = "lambda.min"))
selected_features_2 <- rownames(selected_coefficients_2)[selected_coefficients_2[, 1] != 0][-1] 
print("Selected Features:")
print(selected_features_2)

x_selected_2 <- x_train_2[, selected_features_2]
x_test_selected_2 <- x_test_2[, selected_features_2]

original_vars_2 <- colnames(x_train_2)
removed_features_2 <- setdiff(original_vars_2, selected_features_2)
print("Removed Features:")
print(removed_features_2)
```

```{r}
stan_data_2 <- list(
  N = nrow(train_data),
  J = length(unique(train_data$census_msa)),
  K = ncol(x_selected_2),
  y = y_train,
  x = x_selected_2,
  census = as.numeric(as.factor(train_data$census_msa))
)

stan_model_census <- "
data {
  int<lower=0> N;
  int<lower=0> J;
  int<lower=0> K;
  int<lower=0,upper=1> y[N];
  matrix[N, K] x;
  int<lower=1,upper=J> census[N];   // census_msa categories
}

parameters {
  vector[K] beta;
  vector[J] alpha;   // Group-level intercepts for census_msa
  real mu_alpha;
  real<lower=0> sigma_alpha;
}

model {
  alpha ~ normal(mu_alpha, sigma_alpha);
  beta ~ normal(0, 1);
  mu_alpha ~ normal(0, 1);
  sigma_alpha ~ cauchy(0, 1.5);
  
  y ~ bernoulli_logit(alpha[census] + x * beta);
}

generated quantities {
  vector[N] theta;
  vector[N] log_lik;
  for (n in 1:N) {
    theta[n] = inv_logit(alpha[census[n]] + x[n] * beta);
    log_lik[n] = bernoulli_lpmf(y[n] | theta[n]); 
  }
}
"

stan_model_census <- stan_model(model_code = stan_model_census)
fit_final_2 <- sampling(stan_model_census, data = stan_data_2, iter = 2000, chains = 4, seed = 123, cores = 4)
```

```{r}
print(fit_final_2, pars = c("beta", "alpha", "mu_alpha", "sigma_alpha"))
```

# Use of loo package for comparison of the two stan models

```{r}
library("loo")
log_lik_qualification <- extract_log_lik(fit_final, parameter_name = "log_lik")
loo_qualification <- loo(log_lik_qualification)
log_lik_census <- extract_log_lik(fit_final_2, parameter_name = "log_lik")
loo_census <- loo(log_lik_census)
comparison <- loo_compare(loo_qualification, loo_census)
print(comparison)
```

# Posterior distribution

```{r}
library(bayesplot)

posterior_samples <- extract(fit_final_2)
all_samples <- cbind(
  as.matrix(posterior_samples$beta),       
  as.matrix(posterior_samples$alpha),      
  posterior_samples$mu_alpha,              
  posterior_samples$sigma_alpha            
)
colnames(all_samples) <- c(
  paste0("beta[", 1:ncol(posterior_samples$beta), "]"),
  paste0("alpha[", 1:ncol(posterior_samples$alpha), "]"),
  "mu_alpha", 
  "sigma_alpha"
)

mcmc_areas(
  as.matrix(all_samples),       
  pars = colnames(all_samples),   
  prob = 0.95                    
)

```

```{r}
selected_feature_names <- colnames(x_selected_2)
beta_mapping <- data.frame(
  Beta = paste0("beta[", seq_along(selected_feature_names), "]"),  
  Predictor = selected_feature_names                            
)

beta_mapping
```


# Make predictions on test data using the posterior samples from the model of census_msa

```{r}
posterior_alpha <- extract(fit_final_2, pars = "alpha")$alpha
posterior_beta <- extract(fit_final_2, pars = "beta")$beta

stan_data_test <- list(
  N = nrow(test_data),   # Total number of test observations
  J = length(unique(train_data$census)),   # Number of unique census groups in the training data
  K = ncol(x_test_selected_2),   # Number of predictors included in the test matrix
  x = x_test_selected_2,   # The matrix of selected predictors for the test data
  census = as.numeric(as.factor(test_data$census_msa))   # Encoded group IDs for the census_msa variable in the test data
)

# Compute the posterior linear predictor for each test observation
linear_predictors <- sapply(1:nrow(x_test_selected_2), function(n) {
  posterior_alpha[, stan_data_test$census[n]] + 
    as.numeric(matrix(x_test_selected_2[n, ], nrow = 1) %*% t(posterior_beta))
})

# Convert the linear predictors into probabilities using the sigmoid function
predicted_probabilities <- colMeans(1 / (1 + exp(-linear_predictors)))
test_data$predicted_probabilities <- predicted_probabilities
```


# Grouped prediction analysis

```{r}
grouped_predictions <- test_data %>% group_by(census_msa) %>%
  summarize(
    mean_predicted = mean(predicted_probabilities),
    mean_observed = mean(h1n1_vaccine)
  )

print(grouped_predictions)

ggplot(grouped_predictions, aes(x = census_msa, y = mean_predicted)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_point(aes(y = mean_observed), color = "red", size = 3) +
  labs(title = "Predicted vs Observed Vaccination Rates by Census MSA",
       x = "Census MSA",
       y = "Probability") +
  theme_minimal()
```
